Updated Constraints on Self-Interacting Dark Matter from Supernova 1987A

### 2017-06-19 draft.tex
Removed stray EFT reference.

We revisit SN1987A constraints on light, hidden sector gauge bosons (``dark photon'') that are coupled to the standard model through kinetic mixing with the photon. These constraints are realized because excessive bremsstrahlung radiation of the dark photon can lead to rapid cooling of the SN1987A progenitor core, in contradiction to the observed neutrinos from that event. The models we consider are of interest as phenomenological models of strongly self-interacting dark matter (SIDM). We clarify several possible ambiguities in the literature and identify errors in prior analyses. We find constraints on the dark photon mixing parameter that are in rough agreement with the early estimates of Dent et al. \cite{dent_etal12}, but only because significant errors in their analyses fortuitously canceled. Our constraints are in good agreement with subsequent analyses by Rrapaj \& Reddy \cite{rrapaj_reddy16} and Hardy \& Lasenby \cite{hardy_lasenby17}. We estimate the dark photon bremsstrahlung rate using one-pion exchange (OPE), while Rrapaj \& Reddy use a soft radiation approximation (SRA) to exploit measured nuclear scattering cross sections. We find that the differences between mixing parameter constraints obtained through the OPE approximation or the SRA approximation are roughly a factor of $\sim 2-3$. Hardy \& Laseby \cite{hardy_lasenby17} include plasma effects in their calculations finding significantly weaker constraints on dark photon mixing for dark photon masses below $\sim 10\, \mathrm{MeV}$. We do not consider plasma effects. Lastly, we point out that the properties of the SN1987A progenitor core remain somewhat uncertain and that this uncertainty alone causes uncertainty of at least a factor or $\sim 2-3$ in the excluded values of the dark photon mixing parameter. Further refinement of these estimates is unwarranted until either the interior of the SN1987A progenitor is more well understood or additional, large, and heretofore neglected effects, such as the plasma interactions studied by Hardy \& Lasenby \cite{hardy_lasenby17}, are identified.

\end{abstract}

@@

We initiated our study because of a number of ambiguities appearing in the previous literature on this subject. 
In particular, we could not reproduce the constraints of Ref.~\cite{dent_etal12}. During the course of our 
study, we identified a number of errors in the analysis of Ref.~\cite{dent_etal12}. First, it is straightforward 
to demonstrate that the kinematical relationships given in Appendix A of Ref.~\cite{dent_etal12} are 
incorrect. Second, the squared matrix elements given in Eq.~(A3) and Eq.~(C18) of Ref.~\cite{dent_etal12} must 
be incorrect. These matrix elements do not obey the correct symmetries under interchange of incoming 
and/or outgoing momenta. Furthermore, Ref.~\cite{dent_etal12} neglects the mass of the gauge boson, which 
is legitimate in the classic case of the $\sim$meV-mass axion, but not in the present context. Finally, Ref.~\cite{dent_etal12} 
employed an inconsistent model for the permitted energy loss rate from the supernova interior. Our work amounts 
primarily to repeating the calculation of Ref.~\cite{dent_etal12} in order to rectify these oversights. Our primary 
calculation treats nucleon scattering via one-pion exchange (OPE). We have repeated this analysis 
using an effective field theory for the nucleon interaction and find very similar results so we do not 
report on those in detail.

@@

Finally, we note that the interior temperature and density within the progenitor of SN1987A remain 
uncertain at least at the level of tens of percent \citep[e.g.][]{perego15,farmer16}. This is well worth 
noting because the constraints one derives using the techniques in this work or similar techniques 
are quite sensitive to the properties of the SN1987A core, particularly the core temperature. %Though 
%we do not show this result explicitly \akl{Seems to me we should.} 
As shown in Fig. ~\ref{fig:temperature}, we have repeated a portion of our calculations after shifting the core temperature by $\pm 5\, \mathrm{MeV}$ and found that the lower boundary of the excluded region shifts by a factor of $\sim 2$ at low 
mass ($m_{\mathrm{A}'} \lesssim T$ where $T = 30\, \mathrm{MeV}$ in our fiducial calculation, following 
work of previous authors) and considerably more at high mass. 
We conclude that the differences among theoretical techniques for 
computing the excluded region (OPE, SRA, ...) lead to uncertainties that are of the same order or 
smaller than the uncertainty in the excluded region induced by our limited knowledge of the 
properties of the SN1987A progenitor interior. The excluded regions we and other authors delineate 
should be considered to be uncertain by a factor of $\sim 3$. Refining the theoretical approach to this 
problem is therefore unwarranted until a breakthrough in our understanding of the interior of the 
SN1987A progenitor is achieved.

>>>>

We revisit SN1987A constraints on light, hidden sector gauge bosons (``dark photon'') that are coupled to the standard model through kinetic mixing with the photon. These constraints are realized because excessive bremsstrahlung radiation of the dark photon can lead to rapid cooling of the SN1987A progenitor core, in contradiction to the observed neutrinos from that event. The models we consider are of interest as phenomenological models of strongly self-interacting dark matter (SIDM). We clarify several possible ambiguities in the literature and identify errors in prior analyses. We find constraints on the dark photon mixing parameter that are in rough agreement with the early estimates of Dent et al. \cite{dent_etal12}, but only because significant errors in their analyses fortuitously canceled. Our constraints are in good agreement with subsequent analyses by Rrapaj \& Reddy \cite{rrapaj_reddy16} and Hardy \& Lasenby \cite{hardy_lasenby17}. We estimate the dark photon bremsstrahlung rate using one-pion exchange (OPE), while Rrapaj \& Reddy use a soft radiation approximation (SRA) to exploit measured nuclear scattering cross sections. We find that the differences between mixing parameter constraints obtained through the OPE approximation or the SRA approximation are roughly a factor of $\sim 2-3$. Hardy \& Laseby \cite{hardy_lasenby17} include plasma effects in their calculations finding significantly weaker constraints on dark photon mixing for dark photon masses below $\sim 10\, \mathrm{MeV}$. We do not consider plasma effects. Lastly, we point out that the properties of the SN1987A progenitor core remain somewhat uncertain and that this uncertainty alone causes uncertainty of at least a factor of $\sim 2-3$ in the excluded values of the dark photon mixing parameter. Further refinement of these estimates is unwarranted until either the interior of the SN1987A progenitor is more well understood or additional, large, and heretofore neglected effects, such as the plasma interactions studied by Hardy \& Lasenby \cite{hardy_lasenby17}, are identified.
\end{abstract}

@@

We initiated our study because of a number of ambiguities appearing in the previous literature on this subject. 
In particular, we could not reproduce the constraints of Ref.~\cite{dent_etal12}. During the course of our 
study, we identified a number of errors in the analysis of Ref.~\cite{dent_etal12}. First, it is straightforward 
to demonstrate that the kinematical relationships given in Appendix A of Ref.~\cite{dent_etal12} are 
incorrect. Second, the squared matrix elements given in Eq.~(A3) and Eq.~(C18) of Ref.~\cite{dent_etal12} must 
be incorrect. These matrix elements do not obey the correct symmetries under interchange of incoming 
and/or outgoing momenta. Furthermore, Ref.~\cite{dent_etal12} neglects the mass of the gauge boson, which 
is legitimate in the classic case of the $\sim$meV-mass axion, but not in the present context. Finally, Ref.~\cite{dent_etal12} 
employed an inconsistent model for the permitted energy loss rate from the supernova interior. Our work amounts 
primarily to repeating the calculation of Ref.~\cite{dent_etal12} in order to rectify these oversights. Our primary 
calculation treats nucleon scattering via one-pion exchange (OPE).

@@

Finally, we note that the interior temperature and density within the progenitor of SN1987A remain 
uncertain at least at the level of several tens of percent \citep[e.g.][]{perego15,farmer16}. This is well worth 
noting because the constraints one derives using the techniques in this work or similar techniques 
are quite sensitive to the properties of the SN1987A core, particularly the core temperature. 
As an example of this, we have repeated a portion of our calculations after 
shifting the core temperature by $\pm 5\, \mathrm{MeV}$. This result is shown in Figure~\ref{fig:temperature}.  
We find that the lower boundary of the excluded region shifts by a factor of $\sim 2$ at low 
mass ($m_{\mathrm{A}'} \lesssim T$ where $T = 30\, \mathrm{MeV}$ in our fiducial calculation, following 
work of previous authors) and considerably more at high mass. We conclude that the differences among theoretical 
techniques for computing the excluded region (OPE, SRA, ...) lead to uncertainties that are of the same order or 
smaller than the uncertainty in the excluded region induced by our limited knowledge of the 
properties of the SN1987A progenitor interior. The excluded regions we and other authors delineate 
should be considered to be uncertain by a factor of $\sim 3$. Refining the theoretical approach to this 
problem is therefore unwarranted until a breakthrough in our understanding of the interior of the 
SN1987A progenitor is achieved.
